### 2020-06-28 paper/sex_selection_spacing-ver4.tex
Highlight what question is answered by section
Also edited Introduction to avoid duplication and to cut text from
other sections

First, did birth spacing changes bias the fertility estimates for India?
The tempo effect from longer birth intervals makes the total fertility rate a downward 
biased estimate of cohort fertility. \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
Hence, if birth intervals increased substantially over time, India's fertility may 
be higher than generally accepted.

Second, how do the changes in birth spacing and the increased use of abortions impact 
children's chances of survival?
In India, birth intervals have traditionally been shorter with fewer sons, contributing 
to the higher mortality risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Longer birth spacing may, therefore, lower mortality 
\citep{Conde-Agudelo2012,Molitoris2019}.
A counteracting effect is, however, possible if the longer spacing arises because of 
multiple abortions. 
Nevertheless, there has been no attempt to examine how infant mortality changed with 
the increase in sex selection.

@@

To provide a conceptual framework for understanding how birth spacing
may respond to the substantial changes in income, female education, and
female labor force participation that India has experienced, I discuss
in this section relevant theories and prior findings on birth spacing.

@@

\section{How Birth Spacing has Changed\label{sec:results}}

% [How to read tables]
I estimate the model for each spell/education/period subsample, the results of which I use 
to predict average birth spacing, sex ratio, and the probability of having a birth in that 
spell.
Figures \ref{fig:bootstrap_low} through \ref{fig:bootstrap_highest}
show these predicted outcomes for the four education levels separated by the area of 
residence.%
\footnote{
For legibility, the graphs do not show standard errors.
The graphed values, with standard errors, are shown in Appendix Tables 
\ref{tab:avg_sex_ratio_low} through \ref{tab:avg_sex_ratio_highest}.
Furthermore, I do not show the results for the fourth spell the best-educated women since 
the low number of births, especially with one or more boys, make the spacing and sex ratio 
results very noisy.
Results are available upon request.
}

@@

Second, even with sex selection and the associated increases in expected spacing, the
intervals for the other sex compositions have increased even more, and the relative 
pattern, therefore, remains more or less the same.
An example is for the second spell for women with eight to eleven
and twelve or more years of education.

@@

Average spacing is a convenient way to understand the overall changes in behavior but may 
hide crucial differences in the distribution of spacing.
For example, if sex selection drives the changes in birth spacing, most of the increase in 
spacing will come from people who have two or more pregnancies rather than one and,
therefore, from changes in the longer intervals.
In contrast, a general increase in spacing would show up as an increase in spacing across 
the board.

@@

A potential effect of the substantial increases in birth spacing is that TFR may 
underestimate cohort fertility, at least temporarily.
I, therefore, compare predicted fertility based on a variation of the TFR and the 
hazard model by urban/rural and education groups in Table \ref{tab:fertility}.

@@

To make the fertility rate comparable to the hazard model, it uses only births up to 
parity four, and is, therefore, not directly comparable to those in the NFHS reports.
Otherwise, I follow the procedure used in the NFHS reports \citep{Croft2018}.
I calculate age-specific fertility rates for five-year age groups using the number of 
births in months 1 to 36 before the survey month and the number of women in each age 
group.

@@

The prior literature shows that shorter birth spacing, especially very short spacing,
is associated with a higher risk of infant mortality.
Hence, the longer spacing, both from the increasing use of sex-selective abortions and
from secular changes, may directly reduce infant mortality.

To the extent that the longer birth intervals arise from multiple abortions,
the short duration between \emph{pregnancies} could, however, increase mortality risk.
This section addresses how infant mortality has changed and whether there is evidence
of a negative impact of sex-selective abortions.

Starting with the sample used for estimating birth spacing, I select children born 
more than 12 months before the survey month.
I restrict the analyses to parities two and three because of the small number
of births and deaths for parity four.
Furthermore, I do not show results for women with twelve or more years of
education for the 1972--84 period because of the low number of women.

@@

An important caveat is that the estimations do not 
address potential selection problems.
For example, if women who have difficulties conceiving or carrying a pregnancy 
to term also have a higher mortality risk for their offspring, a spurious correlation 
between long birth spacing and mortality may arise \citep{Kozuki2013}. 
Unfortunately, methods used previously to address selection, such
as family fixed effects, do not work well when the number of births is
low as it is for better-educated women \citep{Kozuki2013,Molitoris2019}.
However, in prior research, the fixed effects results and linear probability do not 
deviate substantially.

>>>>

First, did changes in birth spacing bias the fertility estimates for India? 
With longer spacing, mothers will be older at each parity, and this tempo effect makes the 
total fertility rate a downward biased estimate of cohort fertility
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}. 
Hence, if birth intervals increased substantially, India's fertility may be higher 
than generally accepted.

Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?
In India, birth intervals have traditionally been shorter with fewer sons, contributing 
to the higher mortality risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Hence, longer birth spacing, whether from sex selection or secular changes, may reduce 
mortality \citep{Conde-Agudelo2012,Molitoris2019}.
However, a counteracting effect is possible if longer birth intervals arise from multiple 
abortions since the short duration between \emph{pregnancies} could increase mortality. 
Nevertheless, there has been no attempt to examine how infant mortality changed with 
the increase in sex selection.

@@

To provide a framework for understanding how birth spacing responded to the advent 
of sex selection and the substantial changes in income, education, and labor force 
participation, I next discuss relevant theories and prior findings on birth spacing.

@@

\section{How Birth Spacing Changed\label{sec:results}}

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant changes, 
especially the spread of sex selection, that India experienced. 
I estimate the competing risk hazard model for each spell/education/period subsample and 
then predict average birth spacing, sex ratio, and the probability of having a birth 
in that spell.

Figures \ref{fig:bootstrap_low} through \ref{fig:bootstrap_highest}
show the predicted outcomes for the four education levels by area of residence.%
\footnote{
For legibility, the graphs do not show standard errors.
The graphed values, with standard errors, are shown in Appendix Tables 
\ref{tab:avg_sex_ratio_low} through \ref{tab:avg_sex_ratio_highest}.
Furthermore, I do not show the results for the fourth spell the best-educated women since 
the low number of births, especially with one or more boys, make the spacing and sex ratio 
results very noisy.
Results are available upon request.
}

@@

[TK this is not clear]
Second, even with sex selection and the associated increases in expected spacing, the
intervals for the other sex compositions have increased even more, and the relative 
pattern, therefore, remains more or less the same.
An example is for the second spell for women with eight to eleven
and twelve or more years of education.

@@

To better understand what factors are behind the changes in average spacing, I next 
examine the distribution of birth spacing---represented by the 25th and 75th percentiles 
of predicted birth intervals. 
This approach can capture differences across groups with similar changes in average spacing. 
For example, imagine that sex selection is behind an increase in average birth spacing, 
but that women still try to conceive early. 
In that case, we should see little change in the 25th percentile, but the 75th percentile 
spacing becomes much longer because those who use sex selection will have two or more 
pregnancies rather than one. 
In contrast, a general increase in spacing would show up as lengthening of both percentiles.

@@

The tempo effect from longer birth intervals means that the total fertility rate may 
underestimate cohort fertility. 
The next question I address is, therefore, to what extent did the changes bias the 
fertility estimates for India? 
To this end, I compare fertility based on a variation of the total fertility rate with 
predicted cohort fertility from the hazard model results.

@@

Table \ref{tab:fertility} shows the two fertility measures by urban/rural and education.
Because the hazard model predictions only use births up to parity four, I calculate
fertility rates using only the same births.
Hence, my fertility rates are not directly comparable to those in the NFHS reports.
Otherwise, I follow the procedure used in the NFHS reports \citep{Croft2018}.
I calculate age-specific fertility rates for five-year age groups using the number of 
births in months 1 to 36 before the survey month and the number of women in each age 
group.

@@

The final question I address is whether there is an association between infant mortality 
and increases in birth spacing and sex selection. 
Starting with the sample used for estimating birth spacing, I select children born more 
than 12 months before the survey month. 
I restrict the analyses to parities two and three because of the small number of births 
and deaths for parity four. 
Furthermore, I do not show the results for women with twelve or more years of education 
for the 1972--84 period because of the small number of women.

@@

An important caveat is that the estimations do not 
address potential selection problems.
For example, if women who have difficulties conceiving or carrying a pregnancy 
to term also have a higher mortality risk for their offspring, a spurious correlation 
between long birth spacing and mortality may arise \citep{Kozuki2013}. 
Unfortunately, methods used previously to address selection, such
as family fixed effects, do not work well when the number of births is
low as it is for better-educated women \citep{Kozuki2013,Molitoris2019}.
However, in prior research, the fixed effects and linear probability results do not 
deviate substantially.
